---
title: "Data Manipulation"
author: "Lin Cao"
date: "2026-09-24"
output: github_document
---

This file is for doing data manipulation.
```{r}
library(tidyverse)
```

```{r}
litters_df = 
  read_csv(
    file = "data/FAS_litters.csv", skip = 0,
    na = c("", "NA", ".")
  )
litters_df = janitor::clean_names(litters_df)
```


```{r}
pups_df = 
  read_csv("data/FAS_pups.csv", skip = 3, na = c("", "NA", "."))
pups_df = janitor::clean_names(pups_df)
```

## `select` some variables (column)

```{r}
select(litters_df, group, litter_number) ##选中group和litter_number

select(litters_df, group:gd_of_birth) ##从group到gd_of_birth

select(litters_df, -group) ##删除group

select(litters_df, starts_with("gd")) ##所有gd开头的

select(litters_df, starts_with("gd"), group)

select(litters_df, GROUP = group, everything()) ##把group改名成GROUP，剩下的列都要

select(litters_df, -starts_with("gd")) ##删除所有gd开头的
```

a couple of alternatives
```{r}
rename(litters_df, GROUP = group)

relocate(litters_df, litter_number, gd0_weight) ##把这两个放到最前面
```

In the pups data, seelct litter number, sex, and PD ears.

```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## `filter` some rows

```{r}
filter(litters_df, gd_of_birth == 19)

filter(litters_df, pups_born_alive <= 5)

filter(litters_df, pups_born_alive != 6)

filter(litters_df, group %in% c("Con7", "Con8")) ##只留下group是con7和con8的行

filter(litters_df, group %in% c("Con7", "Con8"), pups_born_alive < 7)

filter(litters_df, pups_born_alive < 4 | pups_born_alive > 8)

drop_na(litters_df, gd0_weight) ##qu diao que shi zhi 
```

In the pups dataset, keep only rows where PD walk is less than 11 and sex is 2

```{r}
filter(pups_df, pd_walk < 11, sex == 2)
```

## `mutate` to change some columns

let's add and change some columns

```{r}
mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight,
  wt_gain_squared = wt_gain ^ 2,
  group = str_to_lower(group))
```

In the pups, create a variable that subtracts 7 from PD walk

```{r}
mutate(
  pups_df,
  pd_walk_minus_7 = pd_walk - 7
)
```

## `arrange` the data

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, desc(pups_born_alive))
```


```{r}
litters_df = 
  read_csv(
    file = "data/FAS_litters.csv", skip = 0,
    na = c("", "NA", ".")
  )
litters_rename_df = janitor::clean_names(litters_df) 
litters_with_gd_df = select(litters_rename_df, group, starts_with("gd"))
litters_no_na_df =  drop_na(litters_with_gd_df)
litters_wt_gain_df = mutate(litters_no_na_df, wt_gain = gd18_weight - gd0_weight)
```


This is good:
```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  select(group, starts_with("gd")) |>
  drop_na() |>
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
  )
```

Load pups, clean names, drop missing, keep litter number and pd variables, add pd walk - 7
```{r}
pups_df = 
  read_csv("data/FAS_pups.csv", skip = 3, na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  drop_na() |>
  select(litter_number, starts_with("pd")) |>
  mutate(
    pd_walk_minus_7 = pd_walk - 7
  )
  
```

sometime the first thing isn't a dataframe
```{r}
litters_df |>
  lm(gd18_weight ~ gd0_weight, data = _)
```

`select` vs `pull`
```{r}
litters_df |>
  select(group)

litters_df |>
  pull(group)
```

never do `$`
```{r}
litters_df$group
```

pipes and viewing
```{r}
litters_df |>
  filter(group %in% c("Con7", "Con8"))
```





